Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2025-02-03 Manuscript.qmd
Completed first-pass revisions for Ecography

abstract: | 
 Occupancy models.

format:
  html:
    theme: simplex

@@

Background Information section

#### Model form and assumptions

Model form.

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process sub-model describes changes in occupancy over time, while the observational process sub-model describes detectability.](Figures/ModelForm.png){#fig-modelform}

Survey design.

![The sampling design of the standard dynamic occupancy model. During seasons (also called primary occasions) sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations (also called surveys or secondary occasions) are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

Assumptions.

# Review methods

Methods.

@@

Coverage.

![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata. The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}

Stamp collection.

@@

Covariates.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;). We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."

library(tidyverse)
library(gt)

covTable <- read_csv("Figures/CovariateTable.csv")

covTable |>
  mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
                             .default = FALSE)) |>
  arrange(Category, Summary, -Prop_Any) |>
  gt(rowname_col = "Covariate", groupname_col = "Category") |>
  tab_options(
    row_group.as_column = TRUE
  ) |>
  tab_row_group(
    label = md("Environmental covariates"),
    rows = Category == "E",
    id = "Env"
  ) |>
  tab_row_group(
    label = md("Structural covariates"),
    rows = Category == "S",
    id = "Struct"
  ) |>
  tab_row_group(
    label = md("All covariates"),
    rows = Category == "O",
    id = "All"
  ) |>
  tab_style(
    style = cell_fill(color = "aquamarine4"),
    locations = cells_row_groups(groups = "Env")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "plum4")
    ),
    locations = cells_row_groups(groups = "Struct")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "gray60")
    ),
    locations = cells_row_groups(groups = "All")
  ) |>
  row_group_order(groups = c("Env", "Struct", "All")) |>
  tab_spanner(label = "Percentage of articles with covariate on parameters:",
              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
                          "Colonisation", "Extinction_Persistence",
                          "Detection"),
              id = "inclusions") |>
  tab_spanner(label = "Percentage which are:",
              columns = c(Prop_Dynamic, Prop_Direct),
              id = "covariates") |>
  tab_spanner(label = "Articles representing this covariate with:",
              columns = c(Prop_Nonlinear, Prop_Interact),
              id = "representations") |>
   cols_label(
    Prop_Any = "Any",
    `Initial occupancy` = html("&psi;1"),
    Occupancy = html("&psi;"),
    Colonisation = html("&gamma;"),
    Extinction_Persistence = html("&epsilon;"),
    Detection = html("&rho;"),
    Prop_Dynamic = "Dynamic",
    Prop_Direct = "Directly observed",
    Prop_Nonlinear = "Non-linear relationships",
    Prop_Interact = "Interactions between covariates"
  ) |>
  text_case_match(
    "Environmental" ~ html("<b>Any Environmental</b>"),
    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),
    "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
    "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
    "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
    "Structural" ~ html("<b>Any Structural</b>"),
    "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),
    "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
    "SPEC" ~ html("<b>Species effect</b>\n<i>Species-level effects</i>"),
    "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),
    "Overall" ~ html("<b>All covariates</b>"),
    .locations = cells_stub()
  ) |>
  cols_hide(c(Category, Summary)) |>
  fmt_percent(decimals = 0) |>
  sub_missing(missing_text = "-") |>
  data_color(
    columns = where(is.numeric),
    method = "numeric",
    palette = "viridis",
    alpha = 0.5
  ) |>
  tab_style(locations = cells_stub(rows = "Environmental"),
            style = cell_fill(color = "aquamarine4", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Environmental"),
            style = list(cell_fill(color = "aquamarine4", alpha = 0.5),
                         cell_text(weight = "bold", style = "italic",
                                   color = "black"))) |>
  tab_style(locations = cells_stub(rows = "Structural"),
            style = cell_fill(color = "plum4", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Structural"),
            style = list(cell_fill(color = "plum4", alpha = 0.5),
                         cell_text(weight = "bold", style = "italic",
                                   color = "black"))) |>
  tab_style(locations = cells_stub(rows = "Overall"),
            style = cell_fill(color = "gray60", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Overall"),
            style = list(cell_fill(color = "gray60", alpha = 0.5),
                         cell_text(weight = "bold",
                                   color = "black"))) |>
  tab_style(locations = cells_row_groups(),
            style = cell_text(color = "white")) |>
  as_raw_html()
```
:::

More covariates.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

Modelling summary.

```{r Modelling summary table}
#| echo: false
#| message: false
#| label: tbl-modelling
#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations. The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review. The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches. 2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."

modellingSummary <- read_csv("Figures/ModellingSummary.csv")

>>>>

abstract: | 
 Since their introduction over twenty years ago, dynamic occupancy models (DOMs) have developed into a powerful and flexible framework for estimating species occupancy across space and throughout time while accounting for imperfect detection. As their popularity has increased and extensions have broadened their capabilities, DOMs have been applied to increasingly diverse datasets and research objectives in applied ecology. In light of this expansion, it is necessary to reassess common practices in building DOMs to identify potential vulnerabilities in model development and identify priorities for future research. We review how DOMs have been applied since their introduction, examining the contexts of their application and the approaches taken for model building. Our findings indicate that DOMs have been readily applied to datasets of wide-ranging size and spatial and temporal scales, resulting in variation in the definition of site occupancy within studies that leads to challenges in building and interpreting DOMs. Current practices for model selection and evaluation also vary considerably, and there remains little research on the impact of these choices on model performance. Where possible we provide recommendations for current users to develop models appropriate for their research contexts, and where uncertainties remain we highlight key priorities for further research into DOMs to support users in producing the most reliable models possible.

format:
  html:
    theme: simplex
    mainfont: Georgia, serif
  docx: default

@@

A considerable portion of ecological research depends on an ability to accurately estimate where and when species occur on the landscape. Understanding the patterns by which various species are distributed has been a fundamental question since the origins of the field, and the subject continues to be an active area of quantitative research and methods development [@humboldt1849]. This is for good reason — species occupancy is a valuable concept with broad relevance in wildlife management and conservation biology. Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and are vital for assessing conservation status, while identification of the potential drivers of occurrence is a key first step in informing potential management action [@mackenzie2013]. Robust knowledge of a species' occupancy patterns can also aid prediction of where it is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].

Irrespective of its importance, 'occupancy' has proven to be a persistently difficult quantity to estimate in practice. This is largely due to the challenges inherent in modelling complex and dynamic natural systems with incomplete data. It is often impossible to reliably determine whether a species truly occupies a site, and it is now well-established that imperfect detection of organisms results in biased estimates of occupancy [@gu2004; @lahoz-monfort2014]. Despite this fact and the ubiquity of imperfect detection in field data, many models in widespread use make no adjustment for it [@kellner2014]. Research has also indicated that many of these tools, including conventional correlative species distribution models, struggle under non-equilibrium conditions where a species’ occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions, often associated with biological invasions and climate change driven range shifts, are increasingly commonplace in the Anthropocene [@bertelsmeier2013; @lenoir2015]. In recent years, awareness of the obstacles to accurate occupancy estimation and the limitations of many popular methods has encouraged interest in somewhat more process-explicit models better suited to these contexts [@briscoe2019].

Within this category are the 'dynamic occupancy models' (DOMs), first introduced by @mackenzie2003 as an extension to earlier static site occupancy models [@mackenzie2002]. In their simplest form, DOMs estimate the probability of site occupancy over multiple time-steps while correcting for imperfect detection using hierarchical observation data. For newcomers to the models we provide a brief introduction to their form and requirements in [Box @nte-Primer]. By accounting for imperfect detection and explicitly modelling colonisation and extinction to capture changes in site occupancy dynamics, DOMs provide answers to common issues in occupancy estimation and offer important advantages over other common models [@guillera-arroita2017a]. Despite these advantages, they require only reasonably simple-to-collect presence absence/data rather than the detailed demographic or abundance data required by other more process-explicit models [@briscoe2019]. This balance between complexity and feasibility makes DOMs an appealing option for many applied ecologists.

::: {#nte-Primer .callout-note icon="false"}
## An introduction to dynamic occupancy models

DOMs are a hierarchical model which links observed detection data with the underlying latent process of changing site occupancy. To do this, they contain two sub-models: one which describes the ecological process of how sites shift between unoccupied and occupied states over time, and one which describes the observation process which describes whether or not a species is detected during surveys at occupied sites.

To separate these processes DOMs require hierarchical data inputs. Occupancy is estimated at independent sites *(i)* during discrete, time-bound intervals termed primary occasions *(t)*. Within primary occasions*,* multiple observations are conducted at the same sites during secondary occasions *(j).* During each observation, a species is either detected (y~i,t,j~ = 1) or not (y~i,t,j~ = 0). Importantly, it is not necessary for all primary occasions to contain the same number of secondary occasions, and missing observations at sites can be accomodated. While secondary occasions are most often represented as repeat observations, other options including same-visit replicates and spatial replicates may also be used – see @guillera-arroita2017a for further discussion on these methods.

![](images/clipboard-2772050481.png){fig-align="center" width="1476"}

DOMs' occupancy sub-model describes how site occupancy changes across primary occasions. In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0). In the first primary occasion, occupancy is determined by a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$. In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probabilities of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$.

![](images/clipboard-3349624601.png){fig-align="center" width="1476"}

The observation sub-model of DOMs describes how the observed detection data (y) relate to the latent occupancy states (z). During any given secondary occasion *j* during primary occasion *t* at site *i,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species. In each secondary occasion, the probability of observing the target species at a site is given as a Bernoulli trial with detection probability *p,* conditional on the site being occupied in primary occasion *t*, such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$. This formulation assumes that no false positive detections exist; that is, observations are never recorded at truly unoccupied sites.

![](images/clipboard-411045638.png){fig-align="center" width="945"}

In this formulation, DOMs estimate four parameters of interest: Initial occupancy (𝛙~1~), colonisation (𝛄), extinction (𝛆), and detection (*p)*. While we have represented each of these parameters as constants for simplicity, they can also be estimated with respect to covariates to describe variation in the occupancy or detection processes - these are typically incorporated via an logit link function, such that DOMs can be considered four interconnected generalised linear models.

DOMs do make some important assumptions of the data used to fit them: i) sites are closed to changes in true occupancy state within each primary occasion, ii) occupancy at each site is independent of other sites, iii) observations within each primary occasion are independent of each other, iv) no false positive detections occur, and v) no un-modelled heterogeneity exists in occupancy or detection probabilities.

For more detailed instruction on DOMs and related models, consider texts by @mackenzie2006 and @kéry2021.
:::

Over the past twenty-some years, continued research on DOMs (and occupancy models more broadly) has established a powerful and flexible modelling framework. Freely available software tools, including program PRESENCE and the R packages *unmarked* and *ubms* among others, have encouraged uptake among applied ecologists [@hines2006; @kery2016; @kellner2022]. Various extensions to the models have further broadened their capabilities – versions of DOMs are available which can account for false positives [@royle2006; @miller2011], model multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimate occupancy for multiple species [@dorazio2010; @devarajan2020]; @bailey2014 provides an overview of many of these extensions. Even with this expansion, there remains room for further growth in applications of DOMs. One area of interest is in DOMs' ability to make more robust predictions of species distributions, a use-case for which they have only seen limited use despite their potential [@kéry2013; @briscoe2019].

Regardless of use-case, DOMs have always been somewhat challenging models to build, evaluate, and interpret. Even defining the term 'occupancy' poses difficulties – what it represents depends on several factors, including the definition of a site, the duration of a primary occasion, and how these relate to the behaviour of the species of interest [@steenweg2018; @wood2022; @valente2024]. This definition of occupancy should inform the nature of the covariates considered for each modelled parameter, although determining which of these factors to include in a model leads to additional questions. Covariate selection is particularly difficult for hierarchical models; with four parameters capable of being estimated with respect to covariates (initial occupancy, colonisation, extinction, and detection), the number of candidate models can rapidly balloon and render exhaustive comparison impossible. The choice of model selection method, and the approach to assessing the resulting models' performance, are important decisions that any user of DOMs must make with respect to their research context and objectives.

The years since the introduction of DOMs have seen enormous changes in how ecologists conduct their research. The amount of data available for modelling has grown considerably, aided by improvements in data sharing and large-scale monitoring efforts. Increased interest in citizen science efforts has unlocked another source of data with its own set of considerations [@altwegg2019]. Technological advances have facilitated the widespread deployment of autonomous detection methods such as camera traps and acoustic monitors, generating massive quantities of fine-scale observation data suitable for analysis with DOMs [@balantic2019]. At the same time, advances in computing have made methods which may have been prohibitively expensive in the past far more accessible for many. While these advances create exciting new opportunities, they also introduce new challenges.

### Objectives

DOMs are an increasingly popular tool for occupancy estimation with broadening applications in applied ecological research. As model users encounter research questions, data structures, and modelling techniques beyond the context of the models' original development, it is necessary to revisit common approaches for building DOMs to identify any potential vulnerabilities, areas of confusion, and uncertainty in best practises. We conduct an in-depth review of studies using DOMs since their introduction, examining how the context of their application and practices for model building and evaluation have changed over the past two decades. We use our findings to identify the largest challenges in building DOMs and to provide recommendations for current users where supported by the literature. Where uncertainties remain, we highlight key priorities for research to better support ecologists in building robust models suited to their research contexts.

# Methods

Our review was constrained to applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions. To capture only these studies, our inclusion criteria required that articles:

i.  Used non-simulated, field-collected, presence-absence data.

ii. Included data from multiple sites which could exist in at least two states, including an occupied and unoccupied state.

iii. Included data from multiple primary occasions, between which sites change states conditional on the prior occasions occupancy state *and* transition probabilities, such as (but not limited to) colonisation and extinction.

iv. Contained at least one parameter describing the detection process.

A pool of candidate articles was generated using two queries on Web of Science on July 26, 2024. The first query included all articles from 2004-2023 which cited @mackenzie2003. To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. These queries resulted in 1469 articles: 897 retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries. To allow comparison of DOMs through time, we divided all articles across five four-year strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each strata we randomly selected 20 articles for inclusion in the review. Articles that did not meet inclusion criteria were replaced from within their own stratum unless no articles remained. For each article, we documented details of the study's research context and their modelling process.

Aspects of research context including authorship, stated research objectives, study taxa and system, and survey methods used to collect data used for modelling. For all authors, the category (academic, government, non-government organisation, or private company) and host country of all affiliated institutions was recorded. To examine the reasons why authors use DOMs, we allocated each article to one or more research objectives category based on the study's stated aims. The six possible categories included "estimating trends*"*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; "testing relationships", where authors examine predefined hypotheses of relationships between covariates and model parameters; "identifying drivers", where authors more broadly explore factors associated with model parameters; "predicting temporally", where authors predict site occupancy under future conditions, "predicting spatially", where authors predict site occupancy to unsurveyed locations, and "developing methods", where authors introduce, test, or demonstrate aspects of DOMs. To develop an understanding of the scale of studies using DOMs, we recorded the quantity of sites, the number of primary and secondary occasions, and the spatial scale of the study location. Spatial scale was defined as the intended area of inference containing all sites, measured to an order of magnitude to account for uncertainty in reporting. Details on the taxa modelled were also recorded, including the quantity of species modelled, their identities, and their conservation status; taxa were denoted as threatened either where they are listed on the IUCN Red List of Species (as of 2024) or when authors explicitly state that they are threatened. This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern, although we acknowledge that in some cases conservation status may have changed since publication.

A key objective of this review was to understand how users of DOMs have navigated the modelling process, through covariate selection through model evaluation. One aspect of this is the type of covariates incorporated in the models – to this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including a general categorisation, whether they were directly observed or remotely sensed, whether they were static or varied between primary occasions, and how they were represented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). We also closely recorded how authors approached model selection. These procedures were sorted into non-exclusive categories which included "*a priori"*, where only one model was considered; candidate suite, where a predefined set of models was considered; "sequential", where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and "simple precursors", where covariate selection was preceded by tests with a simpler model implementation (such as a linear regression or single season occupancy model). Finally, for each model we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in or out-of-sample data.

@@

92 articles were included for this review, fewer than the 100 possible articles due to a deficit of qualifying papers from the 2004-2007 stratum. Based on the acceptance rates within each stratum and the remaining number of unprocessed articles, an estimated 496 of the 1152 unprocessed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).

![Bars indicate the number of articles included in our review as a proportion of the estimated number of articles fitting DOMs in each year. Estimates were calculated using the number of unprocessed articles remaining in each year multiplied by the acceptance rate of the corresponding stratum: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage width="945"}

### Research contexts

Authors of the articles in our sample represent institutions spanning all major sectors of ecological research. 79% of studies included at least one affiliate of an academic institution, 59% a government affiliate, 28% an NGO affiliate, and 12% an affiliate of a private company. 63% of these studies had cross-sector authorship with affiliations from at least two of these categories. While the institutions producing these articles were located in 27 countries across all continents barring Antarctica, 68% of the articles included at least one author affiliated with an institution in the United States of America.

To understand how authors used their DOMs, we classified each study according to their stated research objectives [@fig-Objectives]. These objective categories are non-exclusive, and 37% of studies pursued multiple objectives simultaneously. In the simplest use-case, 36% of articles used DOMs to monitor trends in occupancy state through time, both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015]. Other studies were more focused on the factors influencing species occupancy, either testing pre-defined hypothetical relationships (44% of articles) or taking a more exploratory approach to identifying possible drivers of occupancy (35%). Many articles in the former category target core conservation priorities for their focal species, such as @olson2005's early study assessing the influence of barred owls (*Strix varia)* on the threatened spotted owl (*Strix occidentalis*); in the latter grouping authors examine the influence of a wider variety of factors, as in @huber2017's study testing the relative influence of dozens of habitat covariates on wood warbler (*Phylloscopus sibilatrix*) occupancy. While only 17% of articles used their DOMs to make predictions, the proportion of articles pursing this objective has increased in recent years. These studies tend to have a strong conservation focus – as an example, @mcgowan2020 projects future occupancy for wetland birds under alternative management scenarios. 22% of papers reviewed focused on methods development, representing continued work in the literature on extending and testing DOMs.

![The percentage of articles which pursue each of six non-exclusive research objectives. Percentages are given for each 4-year stratum, as well as for all articles in the study period.](Figures/ObjectiveTrend.jpeg){#fig-Objectives width="945"}

Reviewed studies used data from a diverse range of study systems and focal species. The locations of study sites included in our sample are globally distributed: while a majority of articles use data collected in the United States of America, representatives are included from all continents except Antarctica (@fig-StudyDetails A). These study sites vary considerably in size, ranging from the hyper-local to the continental scale (@fig-StudyDetails B). The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019; @basset2023]. The temporal scale of studies shows similar variability – time elapsed between the first and last survey ranged from under one month to over forty years, with a median duration of 8 years (@fig-StudyDetails F).

The majority of studies in our sample focus on vertebrate taxa [@fig-StudyDetails], particularly birds and mammals. DOMs have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions, including DOMs used to model decadel changes in lichen occupancy and the spread of chronic wasting disease [@belinchón2017; @cook2022]. Others have used DOMs to model mosquito dynamics, a strongly applied use-case with important human health implications [@mores2020; @padilla-torres2013]. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon, Great Plains stream fishes, and whales [@fisher2014; @falke2012; @pendleton2022]. While most studies fit models for a single taxa, 44% fit models to multiple taxa (@fig-StudyDetails D) . 34% of studies fit independent models to multiple species [@otto2012; @peach2019], wheras 12% used explicitly multi-species extensions of DOMs. These included hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020] as well as explicit interaction models which estimated conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].

@@

The presence/absence data required for DOMs was collected using various survey methods (@fig-StudyDetails E). in our sample, 79% of studies used human observational data, 11% used live-trapping methods, and 14% used detections from camera traps. These human-observed data include the 10% of studies that used data collected by citizen-scientists to build detection histories. Citizen science data ranged from coordinated surveys at backyard bird-feeders to interviews with civilians seeking reports of tiger (*Panthera tigris*) signs [@zuckerberg2011; @warrier2020]. Several studies used particularly unique survey methods to determine the presence of a species at a location. For example, in the only application of DOMs to marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy. In another unique implementation, @marescot2020 fit a unique multi-species model treating poachers as a taxon and using ranger reports to create detection histories. Notably, not all of this data was collected in a hierarchical structure with DOMs in mind. In these cases, authors formatted their data into a hierarchical format *post-hoc* using a variety of methods. Some defined primary occasions as arbitrary discrete time intervals, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which predate the DOM, as in a study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

### Modelling practices

To consider how authors incorporate covariates in their DOMs, we recorded all covariates considered for each model in our sample, without regard for whether they were included in final models. @fig-CovariateSummary presents a taxonomy of these covariates, stating the proportion of studies considering each type of covariate, whether covariate data was directly observed at sites, and how covariates responses were represented in the models. We further delineate these covariates by classifying them two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models – while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies). Often these are included as simple covariates on colonisation or extinction, such as the distance to other sites or landscape habitat connectivity metrics [@duggan2011]. Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process and blend attributes of DOMs with patch occupancy models [@broms2016; @risk2011]. Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species. These covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.

![A summary of the covariates considered for the models in our sample. The first row gives the percentage of studies which considered at least one covariate in a given category. The second gives the percentage of covariates considered which are dynamic and vary between occasions; this category does not include covariates used for initial occupancy probabilities which are static by definition. The third row presents the percentage of covariates which are directly observed at sites, rather than being remotely sensed. The final row gives the percentage of covariates in each category where non-linear responses or interactions with other covariates were considered, rather than only testing linear responses.](Figures/CovariateSummary.png){#fig-CovariateSummary}

Covariate data for studies in our sample was either collected directly by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this of course varied depending on the category of covariate (@fig-CovariateSummary). Directly collected data often represents finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely. An average of 43% percent of environmental factors and 94% of structural factors included in reviewed studies were dynamic covariates which varied through time — this too varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@fig-CovariateSummary).

In the conventional DOMs, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function) [@mackenzie2017]. Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified, although non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates. Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms. Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms.

In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter (@fig-CovariateCount). This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter. Note that this does not represent the number of covariates included in the final model formulation used for inference, but it does represent the upper limit of complexity in candidate models. The median number of covariates varies by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates compared to initial occupancy and detection. The lack of any covariates considered for initial occupancy in 37% of studies is particularly notable. Detection probability is also typically represented with low numbers of environmental covariates @fig-CovariateCount.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-CovariateCount width="945"}

To explore how users of DOMs have approached the modelling process, we documented the methods used for model selection and evaluation in all articles in our sample [@tbl-modelling]. Some of these articles included multiple distinct modelling workflows, resulting in a higher total number of models than articles (n = 102 vs. n = 92). Of these models, 76 were fit by maximum likelihood and 24 were Bayesian implementation. 2 models fell into neither category and used machine learning based methods.

In most cases, the DOMs fit in a study did not use all covariates considered for inclusion. 80% of modelling workflows included some form of model selection approach to identify a final model or set of models to make inference from. Approaches varied between the maximum-likelihood and Bayesian (n = 24) implementations in our sample – where 95% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 33% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. For frequentist models, the most popular and conventional approach to model selection (45% of models) involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. The next most popular method in frequentist studies is to use sequential model selection methods (37% of models), where the structure for each model parameter is fit in sequence. For example, this protocol might first identify the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters are fixed. The remainder of frequentist studies (8%) use a variety of other approaches, such as fitting simpler models like single season occupancy models to identify the most informative covariates to use in a dynamic occupancy model. Across all frequentist implementations, 47% of articles use multi-model inference by model-averaging with AIC weights [@burnham2004].

```{r Table_Modelling}
#| echo: false
#| message: false
#| label: tbl-modelling
#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by maximum likelihood or bayesian implementations. The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review. The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches. 2 models included in the 'Overall' column are neural network based and fall into neither the MLE or Bayesian categories."

library(tidyverse)
library(gt)
modellingSummary <- read_csv("Figures/ModellingSummary.csv") |>
  rename(`Maximum likelihood` = Frequentist)
